---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# rODE

The goal of rODE is to explore R and its S4 classes and their differences with Java and Python while exploring simulations by solving differential equations.

`rODE` is based in the extraordinary physics library for computer simulations OpenSourcePhyisics. Tae a look at http://opensourcephysics.org.

The ODE solvers implemented in R so far:

* Euler
* Euler-Richardson
* RK4
* RK45, Dormand-Prince45
* Verlet


## Installation

You can install rODE from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("AlfonsoRReyes/rODE")
```

## Examples
Example scripts are located under the folder `examples` in the package.
These examples make use of a parent class containing a customized rate calculation as well as the step and startup method. The methods that you would commonly find in the base script or parent class are:

* getRate()
* getState()
* step() or doStep()
* setStepSize()
* initialize(), and
* the constructor

These methods are defined in the virtual classes `ODE` and `ODESolver`. Two other classes that serve as definition classes for the ODE solvers are: `AbstractODESolver` and `ODEAdaptiveSolver`.

For instance, the application `KepplerApp.R` needs the class `Kepler` located in the `Kepler.R` script, which is called with `planet <- Kepler(r, v)`, an `ODE` object. The solver for the same application is `RK45` called with `solver <- RK45(planet)`. Since `RK45` is an ODE solver, the script `RK45.R` will be located in the folder `./R` in the package.


## Vignettes
The vignettes contain examples of the use of the various ODE solvers.
For instance, the notebook `Comparison` and ``Kepler` use the ODE solver `RK45`, `FallingParticle` and `Planet` use `Euler``, `Pendulum` makes use of `EulerRichardson`, `Planet` of `Euler`, `Projectile` and `Reaction` of `RK4`, and `KeplerEnergy` uses the ODE solver `Verlet`.




