---
title: "Falling Particle ODE"
author: "Wolfgang Christian / Alfonso R. Reyes"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Falling Particle ODE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Vignettes are long form documentation commonly included in packages. Because they are part of the distribution of the package, they need to be as compact as possible. The `html_vignette` output type provides a custom style sheet (and tweaks some options) to ensure that the resulting html is as small as possible. The `html_vignette` format:

```{r}
library(rODE)
library(ggplot2)

# load the R class that sets up the solver for this application
# source("./inst/examples/FallingParticleODE.R")

source(paste(system.file("examples", package = "rODE"), "FallingParticleODE.R", sep = "/"))


initial_y <- 10   # initial y position
initial_v <- 0    # initial x position
dt        <- 0.01 # delta time for step
```


## Set up the ODE and the solver

```{r}
ball <- FallingParticleODE(initial_y, initial_v)

solver <- Euler(ball)
solver <- setStepSize(solver, dt)
```


- Never uses retina figures
- Has a smaller default figure size
- Uses a custom CSS stylesheet instead of the default Twitter Bootstrap style


## Start the simulation

```{r}
rowVector <- vector("list")
i <- 1
# stop loop when the ball hits the ground
while (ball@state[1] >= 0) {
    rowVector[[i]] <- list(state1 = ball@state[1], 
                           state2 = ball@state[2], 
                           state3 = ball@state[3])
    solver <- step(solver)
    ball <- solver@ode
    # cat(sprintf("%12f %12f ",  ball@state[1], ball@rate[1] ))
    # cat(sprintf("%12f %12f ",  ball@state[2], ball@rate[2] ))
    # cat(sprintf("%12f %12f\n", ball@state[3], ball@rate[3] ))
    i <- i + 1
}

DT <- data.table::rbindlist(rowVector)
```



## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
ggplot(DT, aes(x = state3, y = state1)) + geom_line(col = "blue")
ggplot(DT, aes(x = state3, y = state2)) + geom_line(col = "red")
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes


Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(DT, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))



```{r eval=FALSE}
# #####################
# FallingParticleODE.R
#

setClass("FallingParticleODE", slots = c(
    g = "numeric"
),
    prototype = prototype(
        g = 9.8
    ),
    contains = c("ODE")
    )


setMethod("initialize", "FallingParticleODE", function(.Object, ...) {
    .Object@state <- vector("numeric", 3)
    return(.Object)
})

setMethod("getState", "FallingParticleODE", function(object, ...) {
    # Gets the state variables.
    return(object@state)
})

setMethod("getRate", "FallingParticleODE", function(object, state, rate, ...) {
    # Gets the rate of change using the argument's state variables.
    object@rate[1] <- state[2]
    object@rate[2] <- - object@g
    object@rate[3] <- 1
    invisible(object)

})

# constructor
FallingParticleODE <- function(y, v) {
    .FallingParticleODE <- new("FallingParticleODE")
    .FallingParticleODE@state[1] <- y
    .FallingParticleODE@state[2] <- v
    .FallingParticleODE@state[3] <- 0
    .FallingParticleODE
}

```

